Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-27 paper-new/categorical-model.tex
more squeezing

An intensional model of gradual typing is defined similarly to an extensional model,
with a few key differences that will be discussed below.
%
The starting point is similar to that of the extensional model: an intensional model
will be given by a diagram in the category of CBPV objects, satisfying
additional properties.
%
This time, however, since we are working intensionally, the semantic denotation of
term precision \emph{is} transitive, so we \emph{do} have a horizontal composition
operation on squares. Compare this to the extensional case, where we could only
compose \emph{relations} horizontally, not squares.
What this means is that we can define a functor for composition of value relations
and squares, and a functor for composition of computation relations and squares.

We can specify this compactly as a category internal to the category of CBPV models
and lax morphisms, where we require that the reflexivity, source, and target morphisms
are strict.
In particular, as in the extensional case, there is a CBPV model of ``objects''
$\mathcal M_f$ and a CBPV model of ``arrows'' $\mathcal M_{sq}$. 
There are CBPV morphisms
$r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$,
just as before.
%
But now, we also have a CBPV morphism $m$ from the pullback 
$\mathcal M_{sq} \times_{s = t} \mathcal M_{sq}$ to $M_{sq}$, i.e., ``composition of arrows".
In particular, this consists of a functor 
$m_{\mathcal{V}} : \vsq \times_{\sv = \tv} \vsq \to \vsq$ for composition of value
relations/squares, and a functor $m_{\mathcal{E}} : \esq \times_{\se = \te} \esq \to \esq$
for composition of computation relations/squares.
Furthermore, $s \circ m = s \circ \pi_1$ and $t \circ m = t \circ \pi_2$.

@@

For the sake of ease of reference, we recap the definition of a step-0 model:
%
\begin{definition}
  A \emph{step-0} model of intensional gradual typing consists of:
  \begin{itemize}
    \item A category internal to the category of CBPV models and lax morphisms,
    where we require that the morphisms $r$, $s$, and $t$ are strict.
    \item A natural transformation $\mho : 1 \Rightarrow U$ such that $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$.
    \item A value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
  \end{itemize}

\end{definition}

@@

We can specify all of this abstractly via categories $\vsim$ and $\esim$ along with
functors $\rvsim : \vf \to \vsim$ and $\svsim, \tvsim : \vsim \to \vf$,
and likewise for computations.
Since bisimilarity of morphisms $f$ and $f'$ requires that they share source and target,
we require that $\svsim$ and $\tvsim$ agree on objects and likewise for $\sesim$ and $\tesim$.
Thus, the objects of $\vsim$ are identified with $\ob(\vf)$.
The morphisms of $\vsim$ are ``bisimilarity proofs'', analogous to the commuting squares of $\vsq$.

There is also a ``symmetry'' endofunctor $\text{sym}_{\mathcal{V}}^\bisim : \vsim \to \vsim$
such that $\svsim \circ \text{sym}_{\mathcal{V}}^\bisim = \tvsim$
and $\tvsim \circ \text{sym}_{\mathcal{V}}^\bisim = \svsim$,
and $\text{sym}_{\mathcal{V}}^\bisim \circ \text{sym}_{\mathcal{V}}^\bisim$ is the identity.
Likewise there is a symmetry endofunctor $\text{sym}_{\mathcal{E}}^\bisim : \esim \to \esim$.

In this setting, we write $\refl_A : A \bisim A$ to mean that $\refl_A \in \ob(\vsim)$,
such that $\svsim(\refl_A) = A = \tvsim(\refl_A)$.
Let $f, f' \in \vf(A_i, A_o)$.
The judgment $\gamma : f \bisim_{A_i, A_o} f'$ is defined to mean:

\begin{enumerate}
  \item $\gamma \in \vsim(\refl_{A_i}, \refl_{A_o})$
  \item $\svsim(\gamma) = f$
  \item $\tvsim(\gamma) = f'$
\end{enumerate}

@@

\begin{definition}\label{def:step-1-model}
A \emph{step-1 intensional model} consists of all the data of a step-0 intensional model along
with:
\begin{itemize}
  \item The categories and functors for bisimiarity described above.
  \item The existence of a distinguished value morphism $\delta_{UB}^* \bisim \id_{UB}$ for each $B$.
  \item A square $\delta_{UB}^* \ltdyn_{Ud}^{Ud} \delta_{UB'}^*$ for all $d : B \rel B'$.
  \item The commutativity condition $U\phi \circ \delta_{UB_1}^* = \delta_{UB_2} \circ U\phi$ for any $\phi \in \ef(B, B')$.
\end{itemize}
% We also require the existence of a distinguised computation morphism $\delta_{FA}^* \bisim \id_{FA}$ for each $A$,
% and a square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$ for all $c : A \rel A'$.
\end{definition}

@@

The perturbations must be preserved by $\times$ $\timesk$, $\arr$, $\tok$, $U$, $\Uk$, $F$, and $\Fk$.

Perturbations must also satisfy a property that we call the ``push-pull'' property,
which is formulated as follows. Let $c : A \rel A'$.
Given a perturbation $\delta \in P_A$, there is a corresponding perturbation
$\push_c(\delta) \in P_{A'}$. % making the following square commute:
%
Likewise, given $\delta' \in P_{A'}$ there is a perturbation $\pull_c(\delta') \in P_A$.
% making the following square commute:

Moreover, push-pull states that the following squares must commute:

\begin{center}
  \begin{tabular}{ m{9em} m{9em} } 
    \begin{tikzcd}[ampersand replacement=\&]
      A \& {A'} \\
      A \& {A'}
      \arrow["\delta"', from=1-1, to=2-1]
      \arrow["{\push_c(\delta)}", from=1-2, to=2-2]
      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
    \end{tikzcd}
    &
    \begin{tikzcd}[ampersand replacement=\&]
      A \& {A'} \\
      A \& {A'}
      \arrow["{\pull_c(\delta')}"', from=1-1, to=2-1]
      \arrow["{\delta'}", from=1-2, to=2-2]
      \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
      \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
    \end{tikzcd}
  \end{tabular}
\end{center}

The analogous property should also hold for computation relations and perturbations.

This is summarized below:

\begin{definition}\label{def:step-2-model}
  A \emph{step-2} model of intensional gradual typing consists of all the data of a step-1 model plus:
  \begin{enumerate}
    \item For each value type $A$, there is a monoid $P_A$ and a homomorphism of monoids
    $\ptb_A : P_A \to \{ f \in \vf(A,A) \mid f \bisim \id \}$.
    \item For each computation type $B$, there is a monoid $\pe_B$ and a homomorphism of monoids
    $\ptb_B : P_B \to \{ g \in \ef(B,B) \mid g \bisim \id \}$.
    \item For all $B$, the distinguished endomorphism $\delta_{UB}^*$ is in $P_{UB}$.
    % \item $\pv_A$ and a monoid homomorphism 
    %   \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]
    % \item $\pe_B$ and a monoid homomorphism
    %   \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]
    \item The functors $\times, \timesk$, $\arr, \tok$, $U, \Uk$, $F, \Fk$ preserve perturbations.
    \item The push-pull property holds for all $c : A \rel A'$ and all $d : B \rel B'$.
  \end{enumerate}
\end{definition}

@@

First, let $\mathcal M$ be any double category with a notion of perturbations,
i.e., for any object $X$ in $\mathcal M$ there is a monoid $P_X$ with a monoid
homomorphism into the endomorphisms on $X$.

\begin{definition}\label{def:quasi-left-representable}
  Let $R$ be a horizontal morphism in $\mathcal M$ between objects $X$ and $Y$.
  We say that $R$ is quasi-left-representable by a vertical morphism $f$ in
  $\mathcal M(X, Y)$ if there are perturbations $\delle_R \in P_X$ and
  $\delre_R \in P_Y$ such that the following squares commute:

  \begin{center}
    \begin{tabular}{ m{7em} m{7em} } 
      % UpL
      \begin{tikzcd}[ampersand replacement=\&]
        X \& {Y} \\
        {Y} \& {Y}
        \arrow["f"', from=1-1, to=2-1]
        \arrow["\delta_R^{r,e}", from=1-2, to=2-2]
        \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
        \arrow[from=2-1, to=2-2, Rightarrow, no head]
      \end{tikzcd}
      &
      % UpR
      \begin{tikzcd}[ampersand replacement=\&]
        X \& {X} \\
        {X} \& {Y}
        \arrow["\delta_R^{l,e}"', from=1-1, to=2-1]
        \arrow["f", from=1-2, to=2-2]
        \arrow[from=1-1, to=1-2, Rightarrow, no head]
        \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \end{tikzcd}
    \end{tabular}
  \end{center}

  We call the first square $\upl$ and the second square $\upr$.

\end{definition}

\begin{definition}\label{def:quasi-right-representable}
  Let $R$ be a horizontal morphism between $X$ and $Y$. We say that $R$ is
  \emph{quasi-right-representable by} $f \in \mathcal M(Y, X)$
  if there exist perturbations $\dellp_R \in P_X$ and
  $\delrp_R \in P_Y$ such that the following squares commute:

  \begin{center}
    \begin{tabular}{ m{7em} m{7em} } 
      % DnR
      \begin{tikzcd}[ampersand replacement=\&]
        {X} \& {Y} \\
        {X} \& {X}
        \arrow["\delta_R^{l,p}"', from=1-1, to=2-1]
        \arrow["f", from=1-2, to=2-2]
        \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
        \arrow[from=2-1, to=2-2, Rightarrow, no head]
      \end{tikzcd}
      &
      % DnL
      \begin{tikzcd}[ampersand replacement=\&]
        {Y} \& {Y} \\
        {X} \& {Y}
        \arrow["f"', from=1-1, to=2-1]
        \arrow["\delta_R^{r,p}", from=1-2, to=2-2]
        \arrow[from=1-1, to=1-2, Rightarrow, no head]
        \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \end{tikzcd}
    \end{tabular}
  \end{center}

  We call the first square $\dnr$ and the second square $\dnl$.

  \end{definition}

@@

Lastly, we require that the model satisfy a weak version of functoriality for 
the CBPV connectives $U,F,\times,\to$. 
First, we will need a definition:

\begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are \emph{quasi-order-equivalent},
  written $c \qordeq c'$, if there exist perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and 
  $\delta^r_1, \delta^r_2 \in \pv_{A'}$ such that the following two squares exist:

  \begin{center}
    \begin{tabular}{ m{9em} m{9em} } 
      \begin{tikzcd}[ampersand replacement=\&]
        A \& {A'} \\
        A \& {A'}
        \arrow["\delta^l_1"', from=1-1, to=2-1]
        \arrow["\delta^r_1", from=1-2, to=2-2]
        \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
        \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \end{tikzcd}
      &
      \begin{tikzcd}[ampersand replacement=\&]
        A \& {A'} \\
        A \& {A'}
        \arrow["\delta^l_2"', from=1-1, to=2-1]
        \arrow["\delta^r_2", from=1-2, to=2-2]
        \arrow["c'", "\shortmid"{marking}, no head, from=1-1, to=1-2]
        \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
      \end{tikzcd}
    \end{tabular}
  \end{center}

  We make the analogous definition for computation relations $d, d' : B \rel B'$.
\end{definition}

We require that the CBPV connectives $U,F,\times,\to$ are \emph{quasi-functorial} on relations,
which we specify as follows:
\begin{itemize}
  \item $U(d \comp d') \qordeq U(d)U(d')$
  \item $F(c \comp c'') \qordeq F(c)F(c')$
  \item $(cc') \to (dd') \qordeq (c \to d)(c' \to d')$
  \item $(c_1c_1') \times (c_2c_2') \qordeq (c_1 \times c_2)(c_1'\times c_2')$
\end{itemize}

We summarize the requirements of a step-3 model below:

\begin{definition}\label{def:step-3-model}
  A \emph{step-3 intensional model} consists
  of all the data of a step-2 intensional model, such that additionally:
  \begin{enumerate}
    \item There are functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$ % TODO: image is thin?
    \item Every value edge $c : A \rel A'$ is quasi-left-representable by $\upf(c)$ and
    every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
    \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
  \end{enumerate}
\end{definition}

>>>>

An intensional model of gradual typing is defined similarly to an extensional model,
with a few key differences that will be discussed below.
%
The starting point is similar to that of the extensional model.
This time, however, since we are working intensionally, the semantic denotation of
term precision \emph{is} transitive, so we \emph{do} have a horizontal composition
operation on squares. Compare this to the extensional case, where we could only
compose \emph{relations} horizontally, not squares.
% What this means is that we can define a functor for composition of value relations
% and squares, and a functor for composition of computation relations and squares.
We can specify this compactly as a category internal to the category of CBPV models
and lax morphisms, where we require that the reflexivity, source, and target morphisms
are strict.
In particular, as in the extensional case, there is a CBPV model of ``objects''
$\mathcal M_f$ and a CBPV model of ``arrows'' $\mathcal M_{sq}$. 
There are strict CBPV morphisms
$r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$,
just as before.
%
But now, we also have a CBPV morphism $m$ from the pullback 
$\mathcal M_{sq} \times_{s = t} \mathcal M_{sq}$ to $M_{sq}$, i.e., ``composition of arrows".
In particular, this consists of a functor 
$m_{\mathcal{V}} : \vsq \times_{\sv = \tv} \vsq \to \vsq$ for composition of value
relations/squares, and a functor $m_{\mathcal{E}} : \esq \times_{\se = \te} \esq \to \esq$
for composition of computation relations/squares.
Furthermore, $s \circ m = s \circ \pi_1$ and $t \circ m = t \circ \pi_2$.

@@

% For the sake of ease of reference, we recap the definition of a step-0 model:
% %
% \begin{definition}
%   A \emph{step-0} model of intensional gradual typing consists of:
%   \begin{itemize}
%     \item A category internal to the category of CBPV models and lax morphisms,
%     where we require that the morphisms $r$, $s$, and $t$ are strict.
%     \item A natural transformation $\mho : 1 \Rightarrow U$ such that $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$.
%     \item A value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
%   \end{itemize}

@@

% We can specify all of this abstractly via categories $\vsim$ and $\esim$ along with
% functors $\rvsim : \vf \to \vsim$ and $\svsim, \tvsim : \vsim \to \vf$,
% and likewise for computations.
% Since bisimilarity of morphisms $f$ and $f'$ requires that they share source and target,
% we require that $\svsim$ and $\tvsim$ agree on objects and likewise for $\sesim$ and $\tesim$.
% Thus, the objects of $\vsim$ are identified with $\ob(\vf)$.
% The morphisms of $\vsim$ are ``bisimilarity proofs'', analogous to the commuting squares of $\vsq$.

% There is also a ``symmetry'' endofunctor $\text{sym}_{\mathcal{V}}^\bisim : \vsim \to \vsim$
% such that $\svsim \circ \text{sym}_{\mathcal{V}}^\bisim = \tvsim$
% and $\tvsim \circ \text{sym}_{\mathcal{V}}^\bisim = \svsim$,
% and $\text{sym}_{\mathcal{V}}^\bisim \circ \text{sym}_{\mathcal{V}}^\bisim$ is the identity.
% Likewise there is a symmetry endofunctor $\text{sym}_{\mathcal{E}}^\bisim : \esim \to \esim$.

% In this setting, we write $\refl_A : A \bisim A$ to mean that $\refl_A \in \ob(\vsim)$,
% such that $\svsim(\refl_A) = A = \tvsim(\refl_A)$.
% Let $f, f' \in \vf(A_i, A_o)$.
% The judgment $\gamma : f \bisim_{A_i, A_o} f'$ is defined to mean:

% \begin{enumerate}
%   \item $\gamma \in \vsim(\refl_{A_i}, \refl_{A_o})$
%   \item $\svsim(\gamma) = f$
%   \item $\tvsim(\gamma) = f'$
% \end{enumerate}

@@

\begin{definition}\label{def:step-1-model}
A \emph{step-1 intensional model} consists of all the data of a step-0 intensional model along
with:
\begin{itemize}
  \item The existence of a reflexive, symmetric relation $\bisim$ on the hom-sets $\vf(A,A')$ and $\ef(B,B')$
  such that $\bisim$ preserves composition.
  \item The existence of a distinguished value morphism $\delta_{UB}^* \bisim \id_{UB}$ for each $B$.
  \item A square $\delta_{UB}^* \ltdyn_{Ud}^{Ud} \delta_{UB'}^*$ for all $d : B \rel B'$.
  \item The commutativity condition $U\phi \circ \delta_{UB_1}^* = \delta_{UB_2} \circ U\phi$ for any $\phi \in \ef(B, B')$.
\end{itemize}
% We also require the existence of a distinguised computation morphism $\delta_{FA}^* \bisim \id_{FA}$ for each $A$,
% and a square $\delta_{FA}^* \ltdyn_{Fc}^{Fc} \delta_{FA'}^*$ for all $c : A \rel A'$.
\end{definition}

@@

The perturbations must be preserved by $\times$ $\timesk$, $\arr$, $\tok$, $U$, $\Uk$, $F$, and $\Fk$.
%
Perturbations must also satisfy a property that we call the ``push-pull'' property,
which is formulated as follows. Let $c : A \rel A'$.
Given a perturbation $\delta \in P_A$, there is a corresponding perturbation
$\push_c(\delta) \in P_{A'}$. % making the following square commute:
%
Likewise, given $\delta' \in P_{A'}$ there is a perturbation $\pull_c(\delta') \in P_A$.
% making the following square commute:
%
Moreover, push-pull states that there are squares 
$\delta \ltsq{c}{c} \push_c(\delta)$ and
$\pull_c(\delta') \ltsq{c}{c} \delta'$.

% \begin{center}
%   \begin{tabular}{ m{9em} m{9em} } 
%     \begin{tikzcd}[ampersand replacement=\&]
%       A \& {A'} \\
%       A \& {A'}
%       \arrow["\delta"', from=1-1, to=2-1]
%       \arrow["{\push_c(\delta)}", from=1-2, to=2-2]
%       \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
%       \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
%     \end{tikzcd}
%     &
%     \begin{tikzcd}[ampersand replacement=\&]
%       A \& {A'} \\
%       A \& {A'}
%       \arrow["{\pull_c(\delta')}"', from=1-1, to=2-1]
%       \arrow["{\delta'}", from=1-2, to=2-2]
%       \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
%       \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
%     \end{tikzcd}
%   \end{tabular}
% \end{center}

The analogous property should also hold for computation relations and perturbations.
The requirements are summarized below:
%
\begin{definition}\label{def:step-2-model}
  A \emph{step-2} model of intensional gradual typing consists of all the data of a step-1 model plus:
  \begin{enumerate}
    \item For each value type $A$, there is a monoid $P_A$ and a homomorphism of monoids
    $\ptb_A : P_A \to \{ f \in \vf(A,A) \mid f \bisim \id \}$.
    \item For each computation type $B$, there is a monoid $\pe_B$ and a homomorphism of monoids
    $\ptb_B : P_B \to \{ g \in \ef(B,B) \mid g \bisim \id \}$.
    \item For all $B$, the distinguished endomorphism $\delta_{UB}^*$ is in $P_{UB}$.
    % \item $\pv_A$ and a monoid homomorphism 
    %   \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]
    % \item $\pe_B$ and a monoid homomorphism
    %   \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]
    \item The functors $\times, \timesk$, $\arr, \tok$, $U, \Uk$, $F, \Fk$ preserve perturbations.
    \item The push-pull property holds for all $c : A \rel A'$ and all $d : B \rel B'$.
  \end{enumerate}
\end{definition}

@@

First, let $\mathcal M$ be any double category with a notion of perturbations,
i.e., for any object $X$ in $\mathcal M$ there is a monoid $P_X$ with a monoid
homomorphism into the endomorphisms on $X$.
%
\begin{definition}\label{def:quasi-left-representable}
  Let $R$ be a horizontal morphism in $\mathcal M$ between objects $X$ and $Y$.
  We say that $R$ is quasi-left-representable by a vertical morphism $f$ in
  $\mathcal M(X, Y)$ if there are perturbations $\delle_R \in P_X$ and
  $\delre_R \in P_Y$ such that there is a square
  $\upl : f \ltsq{R}{r(Y)} \delre_R$ and a square 
  $\upr : \delle_R \ltsq{r(X)}{R}$
%
%   \begin{center}
%     \begin{tabular}{ m{7em} m{7em} } 
%       % UpL
%       \begin{tikzcd}[ampersand replacement=\&]
%         X \& {Y} \\
%         {Y} \& {Y}
%         \arrow["f"', from=1-1, to=2-1]
%         \arrow["\delta_R^{r,e}", from=1-2, to=2-2]
%         \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
%         \arrow[from=2-1, to=2-2, Rightarrow, no head]
%       \end{tikzcd}
%       &
%       % UpR
%       \begin{tikzcd}[ampersand replacement=\&]
%         X \& {X} \\
%         {X} \& {Y}
%         \arrow["\delta_R^{l,e}"', from=1-1, to=2-1]
%         \arrow["f", from=1-2, to=2-2]
%         \arrow[from=1-1, to=1-2, Rightarrow, no head]
%         \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
%       \end{tikzcd}
%     \end{tabular}
%   \end{center}
% %
%   We call the first square $\upl$ and the second square $\upr$. 
%
\end{definition}
%
\begin{definition}\label{def:quasi-right-representable}
  Let $R$ be a horizontal morphism between $X$ and $Y$. We say that $R$ is
  \emph{quasi-right-representable by} $f \in \mathcal M(Y, X)$
  if there exist perturbations $\dellp_R \in P_X$ and
  $\delrp_R \in P_Y$ such that we have a square 
  $\dnr : \dellp_R \ltsq{R}{r(X)} f$
  and a square $\dnl : f \ltsq{r(Y)}{R} \delrp_R$.
  %
  % \begin{center}
  %   \begin{tabular}{ m{7em} m{7em} } 
  %     % DnR
  %     \begin{tikzcd}[ampersand replacement=\&]
  %       {X} \& {Y} \\
  %       {X} \& {X}
  %       \arrow["\delta_R^{l,p}"', from=1-1, to=2-1]
  %       \arrow["f", from=1-2, to=2-2]
  %       \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  %       \arrow[from=2-1, to=2-2, Rightarrow, no head]
  %     \end{tikzcd}
  %     &
  %     % DnL
  %     \begin{tikzcd}[ampersand replacement=\&]
  %       {Y} \& {Y} \\
  %       {X} \& {Y}
  %       \arrow["f"', from=1-1, to=2-1]
  %       \arrow["\delta_R^{r,p}", from=1-2, to=2-2]
  %       \arrow[from=1-1, to=1-2, Rightarrow, no head]
  %       \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
  %     \end{tikzcd}
  %   \end{tabular}
  % \end{center}
  % %
  % We call the first square $\dnr$ and the second square $\dnl$.
  %
  \end{definition}

@@

Lastly, we require that the model satisfy a weak version of functoriality for 
the CBPV connectives $U,F,\times,\to$. 
First, we will need a definition:
%
\begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are \emph{quasi-order-equivalent},
  written $c \qordeq c'$, if there exist perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and 
  $\delta^r_1, \delta^r_2 \in \pv_{A'}$ such that there is a square
  $\delta^l_1 \ltsq{c}{c'} \delta^r_1$ and a square $\delta^l_2 \ltsq{c'}{c} \delta^r_2$.
%
  % \begin{center}
  %   \begin{tabular}{ m{9em} m{9em} } 
  %     \begin{tikzcd}[ampersand replacement=\&]
  %       A \& {A'} \\
  %       A \& {A'}
  %       \arrow["\delta^l_1"', from=1-1, to=2-1]
  %       \arrow["\delta^r_1", from=1-2, to=2-2]
  %       \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  %       \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
  %     \end{tikzcd}
  %     &
  %     \begin{tikzcd}[ampersand replacement=\&]
  %       A \& {A'} \\
  %       A \& {A'}
  %       \arrow["\delta^l_2"', from=1-1, to=2-1]
  %       \arrow["\delta^r_2", from=1-2, to=2-2]
  %       \arrow["c'", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  %       \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
  %     \end{tikzcd}
  %   \end{tabular}
  % \end{center}
%
  We make the analogous definition for computation relations $d, d' : B \rel B'$.
\end{definition}
%
We require that the CBPV connectives $U,F,\times,\to$ are \emph{quasi-functorial} on relations,
which we specify as follows:
\begin{itemize}
  \item $U(d \comp d') \qordeq U(d)U(d')$
  \item $F(c \comp c'') \qordeq F(c)F(c')$
  \item $(cc') \to (dd') \qordeq (c \to d)(c' \to d')$
  \item $(c_1c_1') \times (c_2c_2') \qordeq (c_1 \times c_2)(c_1'\times c_2')$
\end{itemize}

We summarize the requirements of a step-3 model below:
%
\begin{definition}\label{def:step-3-model}
  A \emph{step-3 intensional model} consists
  of all the data of a step-2 intensional model, such that additionally:
  \begin{enumerate}
    \item There are functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$ % TODO: image is thin?
    \item Every value edge $c : A \rel A'$ is quasi-left-representable by $\upf(c)$ and
    every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
    \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
  \end{enumerate}
\end{definition}
